### 2021-11-17 documents/sections/verbs.tex
'to bathe'

\subsection{\rc{e-pɨ} \qu{to bathe}}
\label{sec:bathe}
This verb resisted extensions in \PPek \pcref{sec:pekodian} and \akuriyo \pcref{sec:akuriyo}, but is a regularly inflected \gl{s_a_} verb elsewhere.
Verbs for intransitive \qu{to bathe} are usually detransitivized derivations of transitive roots in Cariban languages, which are reflexes of \rc{pɨ}, or \rc{kupi} in some Venezuelan languages \pcref{tab:bathe}.
As we have shown in \cref{sec:pekodian}, \PPek can be reconstructed as having the pair \rc{ipɨ} (\gl{intr}) / \rc{ɨp(ɨ)} (\gl{tr}).
Thus, while \PPek \qu{to bathe (\gl{tr})} has perfectly regular cognates in other languages of the family, intransitive \qu{to bathe} is divergent in this branch, changing \rc{e-} to \rc{i}.
The reasons for this are entirely mysterious; we are not aware of \rc{i-} as a reflex of the detransitivizer in Pekodian, see also \textcite[506]{meira2010origin}.
However, it should be noted that other languages also show unexpected developments in this verb, considering the apparent glide insertion in Waiwaian or the chaotic distribution of \rc{pɨ} and \rc{kupi} in Venezuelan languages.

>>>>

\subsection{\rc{e-pɨ} \qu{to bathe}}
\label{sec:bathe}
This verb resisted the extensions of \PPek \rc{k-} \pcref{sec:pekodian} and, trivially, \akuriyo \obj{k-} \pcref{sec:akuriyo}.
It took on new \gl{1}\gl{s_a_} prefixes in \PTir \parentext{\trio \obj{s-epɨ-}, \akuriyo \obj{t͡ʃe+pɨ-} \parencites[292]{triomeira1999}[87]{gildea1994akuriyo}} and \PWai \parentext{\hixka \obj{k-ewehɨ-}, \waiwai \obj{k-ejeɸu-} \parencites[195]{hixkaryanaderby1985}[166]{waiwaihawkins1998}}.
The fate of its \carijo reflex \obj{ehɨ} \parencites[72]{koch1908hiana} is unknown; we are not aware of a \yukpa cognate.

Verbs for intransitive \qu{to bathe} are usually typical \gl{s_a_} verbs in Cariban languages, derived with a detransitivizer from a transitive root.
These roots are reflexes of \rc{pɨ}, or \rc{kupi} in some Venezuelan languages \pcref{tab:bathe}.
As we have shown in \cref{sec:pekodian}, \PPek can be reconstructed as having the pair \rc{ipɨ} (\gl{intr}) / \rc{ɨp(ɨ)} (\gl{tr}).
Thus, while \PPek \qu{to bathe (\gl{tr})} has perfectly regular cognates in other languages of the family, intransitive \qu{to bathe} is divergent in this branch, changing \rc{e-} to \rc{i}.
The reasons for this are entirely mysterious; reflexes of \rc{i-} are not attested as a reflex of the detransitivizer in Pekodian, see  \textcite[506]{meira2010origin}.
However, it should be noted that other languages also show unexpected developments in this verb, considering the apparent glide insertion in Waiwaian or the distribution of \rc{pɨ} and \rc{kupi} in Venezuelan languages.
